Lattices of acyclic pipe dreams

### 2025-02-18 acyclicFacets/reports/answerReferee.tex

\item \textsl{\color{gray} Page 12 - second to the last line in proposition 4.3: northeast should be southeast.} \\
Thank you for noticing this inconsistency. We have replaced by ``northwest''.

\item \textsl{\color{gray} Page 14 -- line 1: the second ``if'' should be an ``is''.} \\
Thank you. Corrected.

\item \textsl{\color{gray} Page 15 -- cor 4.12: ``patters'' should be ``pattern''.} \\
Thank you. Corrected.

\item \textsl{\color{gray} Page 15 -- section 4.4: this section needs an example.} \\
We have added two examples in Section 4.4. In particular, we have added Figure 10 were we have grouped permutations of~$[e,\omega]$ according to their $\omega$-recoils and pipe dreams of $\Sigma(\omega)$ according to their canopy.
We have also slightly modified the graph~$G(\omega)$ to make these $\omega$-recoil and canopy maps surjective.

@@

\item \textsl{\color{gray} Page 18 - root function: the product notation $\prod Q_X$ needs to be clarified.} \\
We have added ``where~$\prod Q_{[k-1]\smallsetminus I}$ is the product of the letters~$q_i$ for~$i \in [k-1] \smallsetminus I$ computed in the natural order''.

\item \textsl{\color{gray} Page 19 -- line 4 in example 5.5: the c-sorting word notation looks different from the one in line 2, \LaTeX error?} \\
We are not sure we understood your concern here. In the subword complex, we consider the word~$c\omega_\circ(c)$, obtained by concatenation of~$c$ with the $c$-sorting word~$\omega_\circ(c)$. We have added ``and consider the concatenation~$c\omega_\circ(c)$'' to make this clear.

\end{itemize}

{\bf Additional correction.}
Finally, we have also slightly modified the statement and proof of Lemma 4.7.
The statement was missing a $\max(0, \dots)$, and the proof was not considering the case where the new pipe~$\pi(t)$ does not meet the rectangle~$R$.
This is now rectified.
This problem was observed in discussions with Nathan Reading.

### 2025-02-18 acyclicFacets/draft/latticePipeDreams.tex
last changes

\enlargethispage{.6cm}
We now gather some elementary properties of crossings and contacts in pipe dreams that will be needed later to construct pipe dreams from permutations.
For a pipe dream~$P \in \pipeDreams(\omega)$, we call pipe~$j$ the pipe which enters at row~$j$ and exits at column~$\omega^{-1}(j)$.

@@

\begin{proof}
We already proved the equivalence (i) $\Leftrightarrow$ (ii) in \cref{thm:pipeDreamQuotient}.
The equivalence (ii) $\Leftrightarrow$ (iii) is valid for any lattice quotient.
The equivalences (iii) $\Leftrightarrow$ (iv) $\Leftrightarrow$ (v) $\Leftrightarrow$ (vi) follow from the descriptions of \cref{prop:WOIP} of the inversions of the minimum and the non-inversions of the maximum of a weak order interval.
\end{proof}

@@

\begin{definition}
\label{def:recoils}
Consider the graph~$G(\omega)$ with vertex set~$[n]$ and edge set
\[
\set{ij}{i < j \text{ and } j-i, \; \omega^{-1}(i) > \omega^{-1}(j) \le \#\set{k < i}{\omega^{-1}(k) < \omega^{-1}(j)}}.
\]
Let~$\acyclicOrientations(\omega)$ denote the set of acyclic orientations of~$G(\omega)$.
The \defn{$\omega$-recoil congruence} is the equivalence relation on the weak order interval~$[e, \omega]$ whose equivalence classes are the sets of linear extensions of the acyclic orientations of~$G(\omega)$.
The \defn{$\omega$-recoils} of a permutation~$\pi \in [e,\omega]$ is the acyclic orientation~${\recoils{\pi}{\omega} \in \acyclicOrientations(\omega)}$ such that~$\pi$ is a linear extension of~$\recoils{\pi}{\omega}$.
\end{definition}

@@

\begin{proposition}
\label{prop:latticeHomomorphisms}
The maps~$\recoils{\cdot}{\omega}$, $\insertion{\cdot}{\omega}$, and~$\canopy{\cdot}$ define the following commutative diagram of surjective poset homomorphisms:
\[
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=1.2em,column sep=5em,minimum width=2em]
  {
     [e,\omega]  	&								& \acyclicOrientations(\omega)	\\
					& \acyclicPipeDreams(\omega) 	&								\\
  };
  \path[->] (m-1-1) edge node [above] {$\recoils{\cdot}{\omega}$} (m-1-3);
  \path[->>] (m-1-1) edge node [below] {$\insertion{\cdot}{\omega}\qquad$} (m-2-2.west);
  \path[->] (m-2-2.east) edge node [below] {$\quad\canopy{\cdot}$} (m-1-3);
\end{tikzpicture}
\]
\end{proposition}

@@

\begin{example}
\label{exm:canopy}

%
\begin{figure}
	\centerline{\includegraphics[scale=1]{canopy}}
	\caption{The pipe dream congruence (blue bubbles) and the $31542$-recoil congruence (red bubbles) on the weak order interval~$[12345, 31542]$ (left) and the canopy congruence (red bubbles) on the increasing flip graph on acyclic pipe dreams (right). Each blue bubble is contained in a red bubble.}
	\label{fig:canopy}
\end{figure}
\end{example}

@@

\pagebreak
\subsection{Linear extensions of facets}
\label{subsec:linearExtensionsFacets}

@@

\begin{lemma}
\label{lem:linearExtensionsGreedyFacets}
Let $\greedyFacet$ and $\antiGreedyFacet$ be the greedy and antigreedy facets of $\subwordComplex(Q,\omega)$, respectively.
Then 
\begin{enumerate}
\item $e\in \linearExtensions(\greedyFacet)$.
\item $\omega \in \linearExtensions(\antiGreedyFacet)$.
\item If $e\in \linearExtensions(I)$ then $I=\greedyFacet$.
\end{enumerate}
\end{lemma}

@@

\pagebreak
\begin{remark}
\label{rem:linearExtensionsPartitionSubwordComplexA}
Before proving it, we would like to make the following remarks about theorem.
\begin{enumerate}
\item While the sets~$\linearExtensions(I)$ are order convex, they may not be intervals. See \cref{fig:subwordComplex2,fig:subwordComplex4} (the example of \cref{fig:subwordComplex2} is borrowed from~\cite[Figure~9]{PilaudStump-brickPolytope}).
\item The lower set~$\linearExtensions(Q,\omega)$ may have more than one maximal element, and~$\omega$ is not necessarily maximal in~$\linearExtensions(Q,\omega)$. See \cref{fig:subwordComplex5}.
\item Some of the sets~$\linearExtensions(I)$ may not be included in, nor even meet~$[e,\omega]$. See \cref{fig:subwordComplex5}.
\end{enumerate}
%Note that \cref{fig:subwordComplex4,fig:subwordComplex5} represent type~$A$ subword complexes.
%As in~\cite{PilaudPocchiola, PilaudSantos-brickPolytope, PilaudStump-brickPolytope}, we thus use pseudoline arrangements on sorting networks to represent their facets (and we have represented their contact graphs together).
\end{remark}

>>>>

We now gather some elementary properties of crossings and contacts in pipe dreams that will be needed later to construct pipe dreams from permutations.
For a pipe dream~$P \in \pipeDreams(\omega)$, we call pipe~$j$ the pipe which enters at row~$j$ and exits at column~$\omega^{-1}(j)$.

@@

\begin{proof}
We already proved the equivalence (i) $\Leftrightarrow$ (ii) in \cref{thm:pipeDreamQuotient}.
The equivalence \mbox{(ii) $\Leftrightarrow$ (iii)} is valid for any lattice quotient.
The equivalences (iii) $\Leftrightarrow$ (iv) $\Leftrightarrow$ (v) $\Leftrightarrow$ (vi) follow from the descriptions of \cref{prop:WOIP} of the inversions of the minimum and the non-inversions of the maximum of a weak order interval.
\end{proof}

@@

\begin{definition}
\label{def:recoils}
Consider the graph~$G(\omega)$ with vertex set~$[n]$ and edge set
\[
\set{ij}{i < j \text{ and } j-i, \; \omega^{-1}(i) > \omega^{-1}(j) \le \#\set{k < i}{\omega^{-1}(k) < \omega^{-1}(j)}}.
\]
Let~$\acyclicOrientations(\omega)$ denote the set of acyclic orientations of~$G(\omega)$.
%The \defn{$\omega$-recoil congruence} is the equivalence relation on the weak order interval~$[e, \omega]$ whose equivalence classes are the sets of linear extensions of the acyclic orientations of~$G(\omega)$.
The \defn{$\omega$-recoils} of a permutation~$\pi \in [e,\omega]$ is the acyclic orientation~${\recoils{\pi}{\omega} \in \acyclicOrientations(\omega)}$ such that~$\pi$ is a linear extension of~$\recoils{\pi}{\omega}$.
\end{definition}

@@

\begin{proposition}
\label{prop:latticeHomomorphisms}
The maps~$\recoils{\cdot}{\omega}$, $\insertion{\cdot}{\omega}$, and~$\canopy{\cdot}$ define the following commutative diagram of surjective lattice homomorphisms:
\[
\begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=1.2em,column sep=5em,minimum width=2em]
  {
     [e,\omega]  	&								& \acyclicOrientations(\omega)	\\
					& \acyclicPipeDreams(\omega) 	&								\\
  };
  \path[->] (m-1-1) edge node [above] {$\recoils{\cdot}{\omega}$} (m-1-3);
  \path[->>] (m-1-1) edge node [below] {$\insertion{\cdot}{\omega}\qquad$} (m-2-2.west);
  \path[->] (m-2-2.east) edge node [below] {$\quad\canopy{\cdot}$} (m-1-3);
\end{tikzpicture}
\]
\end{proposition}

@@

\begin{example}
For the permutation~$\omega = 1 n \cdots 2$, the graph~$G(\omega)$ has an isolated vertex~$1$ and a path~$2-3-\dots -n$. The orientations of this path can be seen as sign sequences, corresponding to the recoils of permutations or the canopies of binary trees.
More generally, for the permutation~${\omega = 1 \cdots k n \dots (k+1)}$, the graph~$G(\omega)$ consist of all edges~$ij$ for~$k+1 \le i < j \le n$ such that~$|i-j| \le k$, and the $\omega$-recoil map on permutations and canopy on acyclic pipe dreams were already considered in~\cite{Pilaud-brickAlgebra}.
\end{example}

\begin{example}
\label{exm:canopy}
For~$\omega = 31542$, the graph~$G(\omega)$ has a single edge~$4-5$.
We have grouped the permutations of~$[e,\omega]$ according to their $\omega$-recoils, and the pipe dreams of~$\acyclicPipeDreams(\omega)$ according to their canopy in \cref{fig:canopy}.
%
\begin{figure}
	\centerline{\includegraphics[scale=1]{canopy}}
	\caption{The pipe dream congruence (blue bubbles) and the $31542$-recoil congruence (red bubbles) on the weak order interval~$[12345, 31542]$ (left) and the canopy congruence (red bubbles) on the increasing flip graph on acyclic pipe dreams (right). Each blue bubble is contained in a red bubble.}
	\label{fig:canopy}
\end{figure}
\end{example}

@@

\subsection{Linear extensions of facets}
\label{subsec:linearExtensionsFacets}

@@

\pagebreak
\begin{lemma}
\label{lem:linearExtensionsGreedyFacets}
Let $\greedyFacet$ and $\antiGreedyFacet$ be the greedy and antigreedy facets of $\subwordComplex(Q,\omega)$, respectively.
Then 
\begin{enumerate}
\item $e\in \linearExtensions(\greedyFacet)$.
\item $\omega \in \linearExtensions(\antiGreedyFacet)$.
\item If $e\in \linearExtensions(I)$ then $I=\greedyFacet$.
\end{enumerate}
\end{lemma}

@@

\begin{remark}
\label{rem:linearExtensionsPartitionSubwordComplexA}
Before proving it, we would like to make the following remarks about theorem.
\begin{enumerate}
\item While the sets~$\linearExtensions(I)$ are order convex, they may not be intervals. See \cref{fig:subwordComplex2,fig:subwordComplex4} (the example of \cref{fig:subwordComplex2} is borrowed from~\cite[Figure~9]{PilaudStump-brickPolytope}).
\item The lower set~$\linearExtensions(Q,\omega)$ may have more than one maximal element, and~$\omega$ is not necessarily maximal in~$\linearExtensions(Q,\omega)$. See \cref{fig:subwordComplex5}.
\item Some of the sets~$\linearExtensions(I)$ may not be included in, nor even meet~$[e,\omega]$. See \cref{fig:subwordComplex5}.
\end{enumerate}
%Note that \cref{fig:subwordComplex4,fig:subwordComplex5} represent type~$A$ subword complexes.
%As in~\cite{PilaudPocchiola, PilaudSantos-brickPolytope, PilaudStump-brickPolytope}, we thus use pseudoline arrangements on sorting networks to represent their facets (and we have represented their contact graphs together).
\end{remark}
